---
title: "Assignment 06: Data Scraping"
author: "Yiting He"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A06_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Check your working directory

```{r 1. Project Setup, message = FALSE}
#1 
library(tidyverse)
library(rvest)
library(lubridate)

# Check working directory
getwd()


```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2024 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Set the year to 2024
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2024>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2
# Define the URL
url_2024_durham <- "https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2024"

# Read the HTML content (Make sure this is on a new line!)
page_2024_durham <- read_html(url_2024_durham)

```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * Contact Person
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Day Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "Reginald Hicks", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings)".

```{r scrape.the.data}
#3
# Read the HTML content
the_website <- read_html(url_2024_durham)

# Extract all tables from the webpage
tables <- the_website %>% html_elements("table") %>% html_table(fill = TRUE)

# The system information is in the second table
sys_table <- tables[[2]]

# Extract system details from specific cells
water_system_name <- as.character(sys_table[1, 2])
contact_person     <- as.character(sys_table[4, 2])
ownership          <- as.character(sys_table[2, 5])

# Clean the strings by removing leading/trailing whitespace
water_system_name <- trimws(water_system_name)
contact_person     <- trimws(contact_person)
ownership          <- trimws(ownership)

# The water usage data is in the 13th table
MW_ID <- 10
mw_table <- tables[[13]]

# Convert to matrix to extract monthly values by row/column index
mw_mat <- as.matrix(mw_table)

# Extract the 12 monthly Max Day Use values (MGD)
max_day_use_mgd <- c(
  mw_mat[1:4, 3], # Jan, Feb, Mar, Apr
  mw_mat[1:4, 6], # May, Jun, Jul, Aug
  mw_mat[1:4, 9]  # Sep, Oct, Nov, Dec
)

# Convert to character and clean whitespace
max_day_use_mgd <- as.character(max_day_use_mgd)
max_day_use_mgd <- trimws(max_day_use_mgd)

# Output checks
water_system_name
contact_person
ownership
max_day_use_mgd

```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc... 

5. Create a line plot of the maximum daily withdrawals across the months for 2024, making sure, the months are presented in proper sequence.

```{r create.a.dataframe.from.scraped.data}
#4
# Set up basic time variables
Month <- month.abb
Year <- 2024

# Construct the dataframe using scraped variables from Chunk 3
df_2024 <- data.frame(
  Month = Month,
  WaterSystemName = rep(water_system_name, 12),
  ContactPerson   = rep(contact_person, 12),
  Ownership       = rep(ownership, 12),
  MaxDayUse_MGD   = as.numeric(max_day_use_mgd),
  Year = rep(Year, 12)
)

# Convert Month to a factor with levels to ensure correct chronological ordering
df_2024$Month <- factor(df_2024$Month, levels = month.abb)

# Create a proper Date column for plotting
df_2024$Date <- as.Date(paste(df_2024$Year, match(df_2024$Month, month.abb), 1, sep = "-"))

# Display the resulting dataframe
df_2024

#5. Plot the 2024 Water Usage Data
ggplot(df_2024, aes(x = Month, y = MaxDayUse_MGD, group = 1)) +
  geom_line() +
  geom_point() +
  labs(
    title = paste(water_system_name, "2024 Max Daily Water Usage"),
    x = "Month",
    y = "Max Day Use (MGD)"
  ) +
  theme_minimal()

```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function with two inputs, "PWSID" and "year", that:
  - Creates a URL pointing to the LWSP for that PWSID for the given year
  - Creates a website object and scrapes the data from that object (just as you did above)
  - Constructs a dataframe from the scraped data, mostly as you did above, but includes the PWSID and year provided as function inputs in the dataframe. 
  - Returns the dataframe as the function's output


```{r construct.a.scraping.function}
 
#6.
scrape_lwsp <- function(PWSID, year, MW_ID) {
  
  url <- paste0("https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=", PWSID, "&year=", year)
  website <- read_html(url)
  tables <- website %>% html_elements("table") %>% html_table(fill = TRUE)
  
  sys <- tables[[2]]
  water_system_name <- as.character(sys[1, 2])
  contact_person     <- as.character(sys[4, 2])
  ownership          <- as.character(sys[2, 5])
  
  mw <- as.matrix(tables[[MW_ID]])
  max_day_use_mgd <- as.character(c(mw[1:4, 3], mw[1:4, 6], mw[1:4, 9]))
  
  df <- data.frame(
    PWSID = rep(PWSID, 12),
    Year  = rep(year, 12),
    Month = month.abb,
    WaterSystemName = rep(water_system_name, 12),
    ContactPerson   = rep(contact_person, 12),
    Ownership       = rep(ownership, 12),
    MaxDayUse_MGD   = as.numeric(max_day_use_mgd)
  )
  
  df$Date <- as.Date(paste(df$Year, match(df$Month, month.abb), 1, sep = "-"))
  
  return(df)
}
```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2020. Then show the structure of the dataframe with either the `str()` or the `glimpse()` function. 

```{r fetch.and.plot.Durham.2020.data}
#7
# Fetch data for Durham 2020
df_durham_2020 <- scrape_lwsp("03-32-010", 2020, 13)

# Ensure Month factor ordering
df_durham_2020$Month <- factor(df_durham_2020$Month, levels = month.abb)

# Create the plot
ggplot(df_durham_2020, aes(x = Month, y = MaxDayUse_MGD, group = 1)) +
  geom_line() +
  geom_point() +
  labs(
    x = "Month",
    y = "Max Day Use (MGD)",
    title = "Durham 2020 Max Daily Water Usage"
  )

# Check structure
str(df_durham_2020)

```

8. Use the function above to extract data for Cary (PWSID = '03-92-020') in 2020. Combine this data with the Durham data collected above and create a plot that compares Cary's to Durham's water withdrawals.

```{r fetch.and.plot.Cary.2020.data}
#8
# Fetch data for Cary 2020 using the function
df_cary_2020 <- scrape_lwsp("03-92-020", 2020, 13)

# Combine Durham and Cary dataframes together
df_both_2020 <- rbind(df_durham_2020, df_cary_2020)

# Create a 'City' column based on the PWSID to distinguish the two
df_both_2020$City <- ifelse(df_both_2020$PWSID == "03-32-010", "Durham", "Cary")

# Ensure the Month is a factor with correct chronological levels
df_both_2020$Month <- factor(df_both_2020$Month, levels = month.abb)

# Plot the comparison using ggplot2
ggplot(df_both_2020, aes(x = Month, y = MaxDayUse_MGD, color = City, group = City)) +
  geom_line() +
  geom_point() +
  labs(
    x = "Month",
    y = "Max Day Use (MGD)",
    title = "2020 Max Daily Water Usage Comparison: Durham vs Cary"
  )

```


9. Use the code & function you created above to plot Cary's max daily withdrawal by months for the years 2018 thru 2023. Add a smoothed line to the plot (method = 'loess').

> TIP: See Section 3.2 in the "06_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs (where one of the two inputs does not change). Pipe the output of the map2() function to `bind_rows()` to combine the dataframes into a single one, and use that to construct your plot.

```{r fetch.and.plot.Cary.multiyear.data }
#9
# Define the range of years to scrape
years <- 2018:2023
all_data <- NULL

# Use a loop to scrape each year and combine the results
for (y in years) {
  # Note: MW_ID is set to 13 as shown in the screenshot
  one_year <- scrape_lwsp("03-92-020", y, 13)
  all_data <- rbind(all_data, one_year)
}

# Plot the multi-year trend for Cary
ggplot(all_data, aes(x = Date, y = MaxDayUse_MGD)) +
  geom_line(alpha = 0.3) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "loess", se = FALSE) +
  labs(
    title = "Cary Water Usage Trend (2018-2023)",
    x = "Date",
    y = "Max Day Use (MGD)"
  )
```

Question: Just by looking at the plot (i.e. not running statistics), does Cary have a trend in water usage over time?
> Answer:Looking at the plot and the loess smoothed line, there appears to be a slight upward trend in Cary's maximum daily water withdrawals over the 2018-2023 period. While there is significant seasonal fluctuation each year (typically peaking in summer months), the baseline and peak values seem to be gradually increasing over time.
